Analyzing the Effect of Surface in Tennis Grand Slams

### 2018-09-12 contest/surface-type/paper-tennis.Rmd
added references

As to issue (1) quantifying the effect of court surface on players, there has not been much written about with regards to tennis.  There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].   [@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another.  Other sports analyses do take into account surface type such as grass vs. turf in soccer and football.  Results from these studies show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].

@@

HERE
For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense" tests (e.g. these players should win often when contrasting them to opponents) like in [ventura something] and whether these players do have surface apparent effects. START HERE GOAT [@kovalchik2016]

@@

Readers may object that we are looking at difference between grand slams, which each have their own time period, weather conditions, play time conditions, and "home court effects" instead of differences in surfaces alone.  However, (1) grand slam data is the most readily available and most complete which makes it the best choice for modelling, (2) we adjust for these confounders where we can, and (3) analyzing the difference in the grand slams is still useful as they are considered to be the most prestigious events in tennis.

@@

```{r tab-data, include = TRUE, echo = FALSE}
set.seed(19)
s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
rand_inds <- sample(s_id, 5)
cols <- c("tourney_name", "year", "winner_ioc", "w_pointswon", "winner_name", "winner_rank", "l_pointswon", "loser_name")
knitr::kable(gs[rand_inds, cols], format = "latex",
             row.names = FALSE, col.names = c("Tournament", "Year", "W. Country", "W. Points", "Winner", "W. Rank", "L. Points", "Loser"),
             caption = "\\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.",
             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
```

>>>>

As to issue (1) quantifying the effect of court surface on players, there has not been much written about with regards to tennis.  There are materials available in the literature for forecasting the outcome of tennis matches [@klaassen2003; @newton2005; @mchale2011; @kovalchik2016]] or for assessing whether points within a match are independent and identically distributed [@klaassen2001].   [@knottenbelt2012] do take into account surface in their model but do not compare the results of one surface to another.  Other sports analyses do take into account surface type such as grass vs. turf in soccer and football.  Results from these studies show that surface type does have an effect on the game, either directly or indirectly [@andersson2008; @gains2010;].

@@

For issue (2) the player analysis of Nadal, Federer, and Williams, we examine whether our model passes the "common sense" tests  like how the models in [thomas2013] show that commonly well known hockey players also have high status in the model.  We also examine whether these players do have surface apparent effects. Few academic papers have been written about Nadal, Federer, or Williams.  One paper studies Federer's odds of winning when Nadal suddenly withdrew from Wimbledon showed that Federer was too heavily favored by bookmakers [@leitner2009].  One analysis of Williams shows how she has gotten better with age, even past the point when other greats began to decline, but the study does not look at surface type [@five2015].

@@

Readers may object that we are looking at differences between grand slams, which each have their own time period, weather conditions, play time conditions, and "home court effects" instead of differences in surfaces alone.  However, (1) grand slam data is the most readily available and most complete which makes it the best choice at the moment for modelling, (2) we adjust for these confounders where we can, and (3) analyzing the difference in the grand slams is still useful as they are considered to be the most prestigious events in tennis.

@@

```{r tab-data, include = TRUE, echo = FALSE}
w_id <- which(gs$winner_name == "Serena Williams")[1:2]
f_id <- which(gs$winner_name == "Roger Federer")[1:2]
n_id <- which(gs$winner_name == "Rafael Nadal")[1:2]
s_id <- which(gs$winner_name %in% c("Serena Williams", "Roger Federer", "Rafael Nadal"))
cols <- c("winner_name", "tourney_name", "year", "winner_ioc", "w_pointswon", "winner_rank", "l_pointswon", "loser_rank")
knitr::kable(gs[c(w_id, f_id, n_id), cols], format = "latex",
             row.names = FALSE, col.names = c("Winner", "Tournament", "Year", "W. IOC", "W. Points",  "W. Rank", "L. Points", "L. Rank"),
             caption = "\\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.",
             booktabs = TRUE) %>% kable_styling(latex_options = "striped")
```

### 2018-09-12 contest/surface-type/paper-tennis.tex
added references

  \title{Analyzing the Effect of Surface in Tennis Grand Slams}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Kayla Frisoli, Shannon Gallagher, and Amanda Luby}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{September 10, 2018}

@@

As to issue (1) quantifying the effect of court surface on players,
there has not been much written about with regards to tennis. There are
materials available in the literature for forecasting the outcome of
tennis matches (Klaassen and Magnus 2003; Newton and Keller 2005; McHale
and Morton 2011) or for assessing whether points within a match are
independent and identically distributed (Klaassen and Magnus 2001).
(Knottenbelt, Spanias, and Madurska 2012) do take into account surface
in their model but do not compare the results of one surface to another.
Other sports analyses do take into account surface type such as grass
vs.~turf in soccer and football. Results from these studies show that
surface type does have an effect on the game, either directly or
indirectly {[}Andersson, Ekblom, and Krustrup (2008); Gains et al.
(2010);{]}.

@@

HERE For issue (2) the player analysis of Nadal, Federer, and Williams,
we examine whether our model passes the ``common sense'' tests
(e.g.~these players should win often when contrasting them to opponents)
like in {[}ventura something{]} and whether these players do have
surface apparent effects. START HERE GOAT (Kovalchik 2016)

@@

Readers may object that we are looking at difference between grand
slams, which each have their own time period, weather conditions, play
time conditions, and ``home court effects'' instead of differences in
surfaces alone. However, (1) grand slam data is the most readily
available and most complete which makes it the best choice for
modelling, (2) we adjust for these confounders where we can, and (3)
analyzing the difference in the grand slams is still useful as they are
considered to be the most prestigious events in tennis.

@@

\caption{\label{tab:tab-data}\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.}
\centering
\begin{tabular}[t]{lrlrlrrl}
\hiderowcolors
\toprule
Tournament & Year & W. Country & W. Points & Winner & W. Rank & L. Points & Loser\\
\midrule
\showrowcolors
US Open & 2014 & USA & 58 & Serena Williams & 1 & 31 & Taylor Townsend\\
US Open & 2013 & ESP & 136 & Rafael Nadal & 2 & 112 & Philipp Kohlschreiber\\
Australian Open & 2015 & ESP & 91 & Rafael Nadal & 3 & 51 & Mikhail Youzhny\\
US Open & 2013 & USA & 65 & Serena Williams & 1 & 41 & Yaroslava Shvedova\\
Australian Open & 2017 & USA & 73 & Serena Williams & 2 & 60 & Lucie Safarova\\
\bottomrule
\end{tabular}
\end{table}
\rowcolors{2}{white}{white}

@@

@@

>>>>

  \title{Analyzing the Effect of Surface in Tennis Grand Slams}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Kayla Frisoli, Shannon Gallagher, and Amanda Luby}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{September 12, 2018}

@@

As to issue (1) quantifying the effect of court surface on players,
there has not been much written about with regards to tennis. There are
materials available in the literature for forecasting the outcome of
tennis matches (Klaassen and Magnus 2003; Newton and Keller 2005; McHale
and Morton 2011; Kovalchik 2016){]} or for assessing whether points
within a match are independent and identically distributed (Klaassen and
Magnus 2001). (Knottenbelt, Spanias, and Madurska 2012) do take into
account surface in their model but do not compare the results of one
surface to another. Other sports analyses do take into account surface
type such as grass vs.~turf in soccer and football. Results from these
studies show that surface type does have an effect on the game, either
directly or indirectly {[}Andersson, Ekblom, and Krustrup (2008); Gains
et al. (2010);{]}.

@@

For issue (2) the player analysis of Nadal, Federer, and Williams, we
examine whether our model passes the ``common sense'' tests like how the
models in {[}thomas2013{]} show that commonly well known hockey players
also have high status in the model. We also examine whether these
players do have surface apparent effects. Few academic papers have been
written about Nadal, Federer, or Williams. One paper studies Federer's
odds of winning when Nadal suddenly withdrew from Wimbledon showed that
Federer was too heavily favored by bookmakers (Leitner, Zeileis, and
Hornik 2009). One analysis of Williams shows how she has gotten better
with age, even past the point when other greats began to decline, but
the study does not look at surface type (Morris 2015).

@@

Readers may object that we are looking at differences between grand
slams, which each have their own time period, weather conditions, play
time conditions, and ``home court effects'' instead of differences in
surfaces alone. However, (1) grand slam data is the most readily
available and most complete which makes it the best choice at the moment
for modelling, (2) we adjust for these confounders where we can, and (3)
analyzing the difference in the grand slams is still useful as they are
considered to be the most prestigious events in tennis.

@@

\caption{\label{tab:tab-data}\label{tab:data}Example of the grand slam data.  It includes winner and loser attributes, match attributes, and tournament attributes.  Not all attributes are shown here.}
\centering
\begin{tabular}[t]{llrlrrrr}
\hiderowcolors
\toprule
Winner & Tournament & Year & W. IOC & W. Points & W. Rank & L. Points & L. Rank\\
\midrule
\showrowcolors
Serena Williams & Australian Open & 2013 & USA & 52 & 3 & 18 & 110\\
Serena Williams & Australian Open & 2013 & USA & 70 & 3 & 41 & 112\\
Roger Federer & Australian Open & 2013 & SUI & 95 & 2 & 63 & 46\\
Roger Federer & Australian Open & 2013 & SUI & 111 & 2 & 86 & 40\\
Rafael Nadal & Roland Garros & 2013 & ESP & 140 & 4 & 115 & 59\\
Rafael Nadal & Roland Garros & 2013 & ESP & 113 & 4 & 90 & 35\\
\bottomrule
\end{tabular}
\end{table}
\rowcolors{2}{white}{white}

@@

\leavevmode\hypertarget{ref-leitner2009}{}%
Leitner, Christoph, Achim Zeileis, and Kurt Hornik. 2009. ``Is Federer
Stronger in a Tournamentwithout Nadal? An Evaluation of Odds and
Seedings for Wimbledon 2009.'' \emph{Austrian Journal of Statistics} 38
(4): 277--86.

@@

\leavevmode\hypertarget{ref-five2015}{}%
Morris, Benjamin. 2015. ``Serena Williams and the Difference Between
All-Time Great and Greatest of All Time.'' Edited by FiveThirtyEight.
\url{https://fivethirtyeight.com/features/serena-williams-and-the-difference-between-all-time-great-and-greatest-of-all-time/}.
